---
title: "Aiports and Other Travel Terminals with Routes using Leaflet"
author: "Nathan Sowatskey - Coursera/John Hopkins Data Science Specilisation"
always_allow_html: yes
---

```{r setup, echo = FALSE, message = FALSE}
require("leaflet")
require("dplyr")
```
#Introduction

This is an exercise using the [Leaflet JavaScript library](http://leafletjs.com) with R. It is based on airports and airlines data from [OpenFlights.org](http://openflights.org/data.html), as explained below in detail in the "Data" section.

#Data

The data sets that this study is based on are from [OpenFlights.org](http://openflights.org/data.html), specifically in their [GitHub repository](https://github.com/jpatokal/openflights/tree/master/data). Also see this [help](http://openflights.org/help/database.html)

In their, slightly edited, words:

"That GitHub copy is only a sporadically updated static snapshot of the live OpenFlights database. If you would like an up-to-the-minute copy, or you would like your data filtered by any information available to OpenFlights.org (eg. number of routes at the airport), do not hesitate to [contact them](http://openflights.org/about.html)."

For the purposes of this study, I have copied the data available on Wednesday March 1, 2017, 11:30 CET. This is solely so that I have a well-defined baseline. A future version of this study could obtain the data upon demand, or even use other sources of data.

The specific data sets are loaded and illustrated below:

##Routes

See [Route](http://openflights.org/data.html#route) for details.

```{r load_routes}
routes_columns <- c("Airline", "Airline_ID", "Source_airport", "Source_airport_ID", "Destination_airport", "Destination_airport_ID", "Codeshare", "Stops", "Equipment")
routes_df <- read.csv("routes.dat", header = FALSE, col.names = routes_columns, na.strings = "\\N")
departures_per_port <- routes_df %>% group_by(Source_airport) %>% summarise(total_departures = n())
```

##Airports Extended

See [Aiports](http://openflights.org/data.html#airport) for details.

```{r load_airports}
#Set the column names using "Lat" and "Long" to be compatible with Leaflet default naming conventions.
airports_columns <- c("Airport_ID", "Name", "City", "Country", "IATA", "ICAO", "Lat", "Long", "Altitude", "Timezone", "DST", "Tz_database_TZ", "Type", "Source")
#Setting the IATA and ICAO to characters so that we can tidy up missing IATA codes later
airports_colClasses <- c(NA, NA, NA, NA, "character", "character", NA, NA, NA, NA, NA, NA, NA, NA)
airports_df <- read.csv("airports-extended.dat", header = FALSE, col.names = airports_columns, na.strings = "\\N", colClasses = airports_colClasses)
#Remove invalid data
airports_df <- airports_df[(airports_df$Lat != 1 & airports_df$Long != 1),]
airports_df <- airports_df[(airports_df$Lat != 0 & airports_df$Long != 0),]
airports_df <- airports_df[(airports_df$Lat != -1.1111 & airports_df$Long != -1.1111),]
#Add the departures per port value to airports_df
airports_df$total_departures <- departures_per_port[match(airports_df$IATA, departures_per_port$Source_airport), 2]$total_departures
#Set all NAs, which is where there is no departure data, to 0, so that the radius functions work in Leaflet
airports_df$total_departures[is.na(airports_df$total_departures)] <- 0
#Set missing IATA codes to the ICAO value, used for labels in maps, see the airports_colClasses above
airports_df$IATA[is.na(airports_df$IATA)] <- as.character(airports_df$ICAO[is.na(airports_df$IATA)])
#Set still missing IATA codes to name
airports_df$IATA[is.na(airports_df$IATA)] <- as.character(airports_df$Name[is.na(airports_df$IATA)])
#Remove columns that we are not using
airports_df <- subset(airports_df, select=-c(City, Country, ICAO, Altitude, Timezone, DST, Tz_database_TZ, Source))
#Create separate data frames for different types
ports_df <- airports_df[airports_df$Type == "port",] 
stations_df <- airports_df[airports_df$Type == "station",]
unknown_df <- airports_df[airports_df$Type == "unknown",]
airports_df <- airports_df[airports_df$Type == "airport",]
```

#Map with Aiports and Terminals

## Icons Map

```{r icons, echo = FALSE}
airport_icon <- makeIcon(
  iconUrl = "airport_icon.png",
  iconWidth = 225/10, iconHeight = 225/10,
  iconAnchorX = 225/10/2, iconAnchorY = 225/10/2
)

port_icon <- makeIcon(
  iconUrl = "port_icon.jpg",
  iconWidth = 204/10, iconHeight = 204/10,
  iconAnchorX = 204/10/2, iconAnchorY = 204/10/2
)

station_icon <- makeIcon(
  iconUrl = "station_icon.png",
  iconWidth = 225/10, iconHeight = 225/10,
  iconAnchorX = 225/10/2, iconAnchorY = 225/10/2
)

unknown_icon <- makeIcon(
  iconUrl = "unknown_icon.png",
  iconWidth = 225/10, iconHeight = 225/10,
  iconAnchorX = 225/10/2, iconAnchorY = 225/10/2
)

```

This map uses icons to show the different ports types, with the IATA or ICAO code as a label, and the name as a popup. Note that not all ports have IATA codes, so the ICAO code is used instead.

```{r map_with_airports_icons, message = FALSE}
airports_map_icons <- leaflet() %>% addTiles() %>% 
  addMarkers(data = airports_df, popup = ~Name, label = ~IATA, clusterOptions = markerClusterOptions(), icon = airport_icon)  %>% 
  addMarkers(data = ports_df, popup = ~Name, label = ~IATA, clusterOptions = markerClusterOptions(), icon = port_icon)  %>% 
  addMarkers(data = stations_df, popup = ~Name, label = ~IATA, clusterOptions = markerClusterOptions(), icon = station_icon)  %>% 
  addMarkers(data = unknown_df, popup = ~Name, label = ~IATA, clusterOptions = markerClusterOptions(), icon = unknown_icon)
airports_map_icons
```

##Circles Map

This map uses circle markers, with colours to show the different ports types, with the IATA code as a label, and the name as a popup. The circle size represents the number of departures from the port.

At the time of writing, the radius is not being calculated as expected, which is why it is being added to the popup, for debugging purposes.

```{r map_with_airports_circlemarkers, message = FALSE}
airports_map_circles <- leaflet() %>% addTiles() %>% 
  addCircleMarkers(data = ports_df, popup = ~paste(Name, " - ", max(sqrt(total_departures), 10), sep = ""), label = ~IATA, clusterOptions = markerClusterOptions(), color = "blue", radius = ~max(sqrt(total_departures), 10))  %>% 
  addCircleMarkers(data = stations_df, popup = ~paste(Name, " - ", max(sqrt(total_departures), 10), sep = ""), label = ~IATA, clusterOptions = markerClusterOptions(), color = "black", radius = ~max(sqrt(total_departures), 10))  %>% 
  addCircleMarkers(data = unknown_df, popup = ~paste(Name, " - ", max(sqrt(total_departures), 10), sep = ""), label = ~IATA, clusterOptions = markerClusterOptions(), color = "red", radius = ~max(sqrt(total_departures), 10)) %>%
  addCircleMarkers(data = airports_df, popup = ~paste(Name, " - ", max(sqrt(total_departures), 10), sep = ""), label = ~IATA, clusterOptions = markerClusterOptions(), color = "purple", radius = ~max(sqrt(total_departures), 10))  %>%
  addLegend(labels = c("Airport", "Sea Port", "Train Station", "Unknown"), colors = c("purple", "blue", "black", "red"))
airports_map_circles
```

##Icons and Circles

This is a combination of the two ideas, with icon markers and circles.

```{r map_with_airports_icons_and_circles, message = FALSE}
airports_map_icons_and_circles <- leaflet() %>% addTiles() %>% 
  addMarkers(data = airports_df, clusterOptions = markerClusterOptions(), icon = airport_icon)  %>% 
  addMarkers(data = ports_df, clusterOptions = markerClusterOptions(), icon = port_icon)  %>% 
  addMarkers(data = stations_df, clusterOptions = markerClusterOptions(), icon = station_icon)  %>% 
  addMarkers(data = unknown_df, clusterOptions = markerClusterOptions(), icon = unknown_icon) %>%
  addCircleMarkers(data = ports_df, popup = ~paste(Name, " - ", max(sqrt(total_departures), 10), sep = ""), label = ~IATA, clusterOptions = markerClusterOptions(), color = "blue", radius = ~max(sqrt(total_departures), 10))  %>% 
  addCircleMarkers(data = stations_df, popup = ~paste(Name, " - ", max(sqrt(total_departures), 10), sep = ""), label = ~IATA, clusterOptions = markerClusterOptions(), color = "black", radius = ~max(sqrt(total_departures), 10))  %>% 
  addCircleMarkers(data = unknown_df, popup = ~paste(Name, " - ", max(sqrt(total_departures), 10), sep = ""), label = ~IATA, clusterOptions = markerClusterOptions(), color = "red", radius = ~max(sqrt(total_departures), 10)) %>%
  addCircleMarkers(data = airports_df, popup = ~paste(Name, " - ", max(sqrt(total_departures), 10), sep = ""), label = ~IATA, clusterOptions = markerClusterOptions(), color = "purple", radius = ~max(sqrt(total_departures), 10))  %>%
  addLegend(labels = c("Airport", "Sea Port", "Train Station", "Unknown"), colors = c("purple", "blue", "black", "red"))
airports_map_icons_and_circles
```

## Conclusion

For more details about the leaflet package for R 
visit http://rstudio.github.io/leaflet/.
